Estimating Predictive Rate-Distortion Curves via Neural Variational Inference

### 2019-06-12 paper.tex
Updates from Overleaf

@@

As in OCF (Section~\ref{sec:OCF}), we assume that a set of finite sample trajectories $x_{-M} \dots x_{M-1}$ is available, and we aim to compress pasts of length $M$ to predict futures of length $M$. To carry this out, we restrict the PRD objective~(\ref{eq:ib}) to such finite-length pasts and futures: 
\begin{equation}\label{eq:ib-oce-repeat}
	\max_{Z : Z \bot \finitefuture | \finitepast} \left( \operatorname{I}[Z, \finitefuture] - \lambda \cdot \operatorname{I}[\finitepast, Z] \right).
\end{equation}
%where $Z$ ranges over all random variables such that the Markovcondition~(\ref{eq:markov}) holds:
%\begin{equation}
%    Z\bot \finitefuture | \finitepast
%\end{equation}
It will be convenient to equivalently rewrite (\ref{eq:ib-oce-repeat}) as
\begin{equation}\label{eq:prd-loss}
	\min_{Z : Z \bot \finitefuture | \finitepast}	\left[\operatorname{H}[\finitefuture | Z] + \lambda \cdot\operatorname{I}[ \finitepast, Z]\right],
\end{equation}
where $\operatorname{H}[\finitefuture | Z]$ is the \key{prediction loss}. Note that minimizing prediction loss is equivalent to maximizing predictiveness $\operatorname{I}[\finitefuture | Z]$. % is the reduction in prediction loss when predicting

@@

\paragraph{OCF} As discussed in Section~\ref{sec:OCF}, OCF is affected by overfitting, 
and will systematically overestimate the predictiveness achieved at a given rate~\citep{still-optimal-2010}.
To address this problem, we follow the evaluation method used for NPRD, evaluating rate and predictiveness on held-out test data.
We partition the available time series data into a training and test set. We use the training set to create the encoder $P(Z|\finitepast)$ using the Blahut-Arimoto algorithm as described by~\citet{still-optimal-2010}.
We then use the held-out test set to estimate rate and prediction loss.
This method not only enables fair comparison between OCF and NPRD, but also provides a more realistic evaluation, by focusing on the performance of the code $Z$ when deployed on new data samples.
For rate, we use the same variational bound that we use for NPRD, stated in Proposition~\ref{prop:kl-bound}:
\begin{equation}\label{eq:rate-OCF}
\frac{1}{N} \sum_{X_{-M}\dots X_{-1} \in \text{Test Data}} \operatorname{D_{KL}}\infdivx{P(Z|X_{-M}\dots X_{-1})}{s(Z)}
\end{equation}
where $P(Z|X_{-M}\dots X_{-1})$ is the encoder created by the Blahut-Arimoto algorithm, and $s(Z)$ is the marginal distribution of $Z$ on the training set.
$N$ is the number of sample time series in the test data.
In the limit of enough training data, when $s(Z)$ matches the actual population marginal of $Z$, (\ref{eq:rate-OCF}) is an unbiased estimate of the rate. %  If the training set is large enough and 
%\textbf{Is $s$ defined anywhere? same for $r$?}
We estimate the prediction loss on the future observations as the empirical cross-entropy, i.e., the variational bound stated in Proposition~\ref{prop:mi-bound}:
\begin{equation}
\frac{1}{N} \sum_{X_{-M}\dots X_{M-1} \in \text{Test Data}} \E_{Z \sim P(\cdot|\finitepast)} \log P(\finitefuture|Z).
\end{equation}
where $P(\finitefuture|Z)$ is the decoder obtained from the Blahut-Arimoto algorithm on the training set.
Thanks to Propositions~\ref{prop:mi-bound} and \ref{prop:kl-bound}, these quantities provide upper bounds, up to sampling error introduced by finiteness of the held-out data.
Again, sampling error does not bias the results in either direction, unlike overfitting, which introduces a systematic bias.

@@

\begin{figure*}
\centering
\includegraphics[width=0.45\textwidth]{code/figures/foo_pca_2.png}
\includegraphics[width=0.45\textwidth]{code/figures/foo_pca_3.png}
	\caption{Applying PCA to 5000 sampled codes $Z$ for the Random Insertion Process, at $\lambda = 0.6$ (left) and $\lambda = 0.25$ (right). We show the first two principal components. Samples are colored according to the states in the $\epsilon$-machine. There is a small number of samples from sequences that, at $M=15$, cannot be uniquely attributed to any of the states (ambiguous between A and C); these are indicated in black.}\label{fig:latent}
\end{figure*}
% colors = [{"A" : "red", "B" : "green", "C" : "blue", "U" : "black"}[x] for x in states]

>>>>

% NOTES
% - cite this paper https://arxiv.org/pdf/1705.02436.pdf (they have a different IB bound, but also do neural networks)

@@

As in OCF (Section~\ref{sec:OCF}), we assume that a set of finite sample trajectories $x_{-M} \dots x_{M-1}$ is available, and we aim to compress pasts of length $M$ to predict futures of length $M$. To carry this out, we restrict the PRD objective~(\ref{eq:ib}) to such finite-length pasts and futures: 
\begin{equation}\label{eq:ib-oce-repeat}
	\max_{Z : Z \bot \finitefuture | \finitepast} \left( \operatorname{I}[Z, \finitefuture] - \lambda \cdot \operatorname{I}[\finitepast, Z] \right).
\end{equation}
%where $Z$ ranges over all random variables such that the Markovcondition~(\ref{eq:markov}) holds:
%\begin{equation}
%    Z\bot \finitefuture | \finitepast
%\end{equation}
It will be convenient to equivalently rewrite (\ref{eq:ib-oce-repeat}) as
\begin{equation}\label{eq:prd-loss}
	\min_{Z : Z \bot \finitefuture | \finitepast}	\left[\operatorname{H}[\finitefuture | Z] + \lambda \cdot\operatorname{I}[ \finitepast, Z]\right],
\end{equation}
where $\operatorname{H}[\finitefuture | Z]$ is the \key{prediction loss}. Note that minimizing prediction loss is equivalent to maximizing predictiveness $\operatorname{I}[\finitefuture , Z]$. % is the reduction in prediction loss when predicting

@@

\paragraph{OCF} As discussed in Section~\ref{sec:OCF}, OCF is affected by overfitting, 
and will systematically overestimate the predictiveness achieved at a given rate~\citep{still-optimal-2010}.
To address this problem, we follow the evaluation method used for NPRD, evaluating rate and predictiveness on held-out test data.
We partition the available time series data into a training and test set. We use the training set to create the encoder $P(Z|\finitepast)$ using the Blahut-Arimoto algorithm as described by~\citet{still-optimal-2010}.
We then use the held-out test set to estimate rate and prediction loss.
This method not only enables fair comparison between OCF and NPRD, but also provides a more realistic evaluation, by focusing on the performance of the code $Z$ when deployed on new data samples.
For rate, we use the same variational bound that we use for NPRD, stated in Proposition~\ref{prop:kl-bound}:
\begin{equation}\label{eq:rate-OCF}
\frac{1}{N} \sum_{\finitepast \in \text{Test Data}} \operatorname{D_{KL}}\infdivx{P(Z|\finitepast)}{s(Z)}
\end{equation}
where $P(Z|\finitepast)$ is the encoder created by the Blahut-Arimoto algorithm, and $s(Z)$ is the marginal distribution of $Z$ on the training set.
$N$ is the number of sample time series in the test data.
In the limit of enough training data, when $s(Z)$ matches the actual population marginal of $Z$, (\ref{eq:rate-OCF}) is an unbiased estimate of the rate. %  If the training set is large enough and 
%\textbf{Is $s$ defined anywhere? same for $r$?}
We estimate the prediction loss on the future observations as the empirical cross-entropy, i.e., the variational bound stated in Proposition~\ref{prop:mi-bound}:
\begin{equation}
\frac{1}{N} \sum_{\finitepast, \finitefuture \in \text{Test Data}} \E_{Z \sim P(\cdot|\finitepast)} \log P(\finitefuture|Z).
\end{equation}
where $P(\finitefuture|Z)$ is the decoder obtained from the Blahut-Arimoto algorithm on the training set.
Thanks to Propositions~\ref{prop:mi-bound} and \ref{prop:kl-bound}, these quantities provide upper bounds, up to sampling error introduced by finiteness of the held-out data.
Again, sampling error does not bias the results in either direction, unlike overfitting, which introduces a systematic bias.

@@

\begin{figure*}
\centering
\includegraphics[width=0.45\textwidth]{code/figures/foo_pca_2.png}
\includegraphics[width=0.45\textwidth]{code/figures/foo_pca_3.png}
	\caption{Applying Principal Component Analysis to 5000 sampled codes $Z$ for the Random Insertion Process, at $\lambda = 0.6$ (left) and $\lambda = 0.25$ (right). We show the first two principal components. Samples are colored according to the states in the $\epsilon$-machine. There is a small number of samples from sequences that, at $M=15$, cannot be uniquely attributed to any of the states (ambiguous between A and C); these are indicated in black.}\label{fig:latent}
\end{figure*}
% colors = [{"A" : "red", "B" : "green", "C" : "blue", "U" : "black"}[x] for x in states]
